\bf{Inequality in non-cognitive skills: \\ a cross-cohort comparison}

### 2018-05-24 codes/R/slides/cohorts_tapes.Rnw
update slides

@@

\author[Mason]{Orazio Attanasio \and Richard Blundell \\ Gabriella Conti \and Giacomo Mason  }
\institute[UCL Economics]{UCL Department of Economics}

@@

\begin{frame}
\frametitle{Motivation}
\framesubtitle{Inequality in skills}

\bi
\item Socio-emotional skills developed during childhood are important for lifetime outcomes
\item Socioeconomic gradients in these skills are persistent 
\item {\color{red} IMPROVE INTRO...}
\ei

\begin{frame}
\frametitle{This study}
\framesubtitle{Research questions}
\bi
\item Usually, socioemotional constructs are measured using validated scales
    \bi \item e.g. Rutter, BSAG, SDQ \ei
\pause
~\\    
\item \textbf{Q1}: Can we derive \textbf{comparable measures} of socioemotional skills across populations?
    \bi
    \item Most economic literature uses simple sum scores
    \item But does the same instrument measure the same constructs in the same way:
      \bi 
      \item across different populations?
      \item in the same population, at different points in time?
      \ei
    \item And what if only different instruments are available?
    \ei
\pause
~\\    
\item \textbf{Q2}: \textbf{How many latent dimensions} are underlying these instruments?
    \bi
    \item Can they be simply summarised into a single score?
    \ei
\ei
\end{frame}

@@

\bi
\item Focus on socioemotional skills in two British cohorts
  \bi \item British Cohort Study (BCS, b. 1970)
  \item Millennium Cohort Study (MCS, b. 2000/1) \ei
\ei
~\\ \pause
\textbf{Contribution}
\bi
\item Construct a common 11-item scale, starting from different survey instruments
\item Investigate its psychometric properties, and find
  \bnm 
  \item \textbf{\textit{Comparability}} in terms of scale (but not location)
  \item \textbf{\textit{Bidimensionality}} -- Externalising and Internalising
  \enm
\ei
~\\ \pause

\textbf{Key findings}
  \bi
  \item Inequality in non-cognitive skills has increased in 1975-2005
  \item Sharpest change for boys rather than girls
  \item {\color{red} OTHER RESULTS}
  \ei

@@

\bi
\item Four main cohorts:
  \bi
  \item \textit{National Survey of Health and Development} (NSHD) -- b. March 1946
  \item \textit{National Child Development Study} (NCDS) -- b. March 1958
  \item \textit{British Cohort Study} (BCS) -- b. April 1970
  \item \textit{Millennium Cohort Study} (MCS) -- b. 2000-2001
  \ei
~\\
\item Follow 15,000+ individuals from birth onwards
\item Include extensive data on
  \bi
  \item Perinatal period (pregnancy, birth outcomes)
  \item Skill formation -- cognitive, socioemotional
  \item Health and health behaviours
  \item Family background and composition
  \item Schooling and labour market outcomes
  \ei
~\\
\pause
\item Not always consistent
  \bi
  \item Different design, timing, questions, survey instruments...
  \item Harmonisation efforts (CLOSER)
  \ei
\ei

@@

\begin{column}{0.6\textwidth}
\bi
\item Data from 5 year follow-up
\item Sample restricted to England
\item Socioemotional skills: \textbf{Rutter A scale} \citep{Rutter1970}
  \bi
  \item 19 items (descriptions of behaviours) from maternal questionnaire
  \item Options: \textit{Doesn't apply, Applies somewhat, Certainly applies}
  \ei
  \item Additional questions from maternal module
\ei
\end{column}

@@

\bi
\item Data from 5 year follow-up
\item Sample restricted to England
\item Complex survey design: rebalance sample to make representative
\item Socioemotional skills: \textbf{Strengths and Difficulties Questionnaire} (SDQ, \citealp{Goodman1994})
  \bi
  \item ``Updated" and expanded version of the Rutter scale 
  \item 25 items (descriptions of behaviours)
  \item Options: \textit{Not true, Somewhat true, Certainly true}
  \ei
\ei

@@

\section{Harmonised scale}

\subsection{Common items}

@@

\bi
\item Rutter and SDQ are different instruments
\item Focus on subset of items that have similar wording between the two instruments
\item New 11-item subscale
\ei

@@

\begin{frame}
\frametitle{Harmonisation}\label{meantab_f}
\framesubtitle{Choice of common items (females) \hyperlink{meantab_m}{\beamerbutton{Males}}}
\centering
\resizebox{!}{4cm}{
<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
require(xtable)

@@

print(xtable(table,
                    align="L{0cm}C{.04\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                    digits = 1
                    ), 
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
                                    paste(" & 
                                              \\multicolumn{4}{c}{\\textbf{BCS}} &
                                              \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9} 
                                            Item &
                                              Text & Cert. Appl. & Smt. Appl. & Appl. &
                                              Text & Cert. True & Smt. True & True \\\\ \\midrule
                                          "
                                            , sep="")
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = ".8\\textwidth",
             floating = FALSE
)

@
} %scalebox
\end{frame}

\subsection{Exploratory Analysis}

@@

How many dimensions underlie the new 11-item scale?
\bi
\item Rutter A scale (BCS) 
  \bi 
  \item Traditionally conceptualised as bidimensional \citep{Blanden2007}
    \bi
    \item \emph{anti-social} -- or aggressive-hyperactive
    \item \emph{neurotic} -- or anxious-fearful
    \ei
  \item Sometimes the former is split into aggressive-antisocial and hyperkinetic \citep{McGee1985}
  \ei
\item Strengths and difficulties (MCS) 
  \bi
  \item Five subscales (emotional, hyperactivity, prosocial, peer, conduct) \citep{Stone2010}
  \item Evidence that lower-dimensional structures can work \citep{Goodman2010}
  \ei
\ei

@@

On our data:
\bi
\item Different methods for assessing number of factors give varied results \hyperlink{numfac}{\beamerbutton{Det.}}
\item We choose two factors to maximise interpretability
\item In both cohorts, EFA delivers a clear-cut split between two factors
\item Factors can be coherently interpreted as externalising ($\sim$ anti-social) and internalising ($\sim$ neurotic)
\ei 
\vspace{12pt}
\centering
\begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
  & \textbf{{\color{OliveGreen}Externalising}}  &    & \textbf{{\color{Orange}Internalising}}  \\
1 & Restless                & 7  & Worried                \\
2 & Squirmy/fidgety         & 8  & Afraid of new things   \\
3 & Fights/bullies children & 9  & Solitary               \\
4 & Cannot settle           & 10 & Unhappy/tearful        \\
5 & Temper tantrums         & 11 & Head/stomachaches \\
6 & Disobedient             &    &                       
\end{tabular}

@@

\begin{frame}
\frametitle{EFA}\label{meantab_fcc2}
\framesubtitle{Choice of common items (females) \hyperlink{meantab_fcc3}{\beamerbutton{3 factors}}}
\centering
\resizebox{!}{4cm}{
<<MEANTAB_FCC, dependson="MEANTABLE", results='asis'>>=
require(xtable)

@@

print(xtable(table,
                    align="L{0cm}C{.04\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.4\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
                    digits = 1
                    ), 
             size="\\small",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
                                    paste(" & 
                                              \\multicolumn{4}{c}{\\textbf{BCS}} &
                                              \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9} 
                                            Item &
                                              Text & Cert. Appl. & Smt. Appl. & Appl. &
                                              Text & Cert. True & Smt. True & True \\\\ \\midrule
                                          "
                                            , sep="")
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = ".8\\textwidth",
             floating = FALSE
)

@
} %scalebox
\end{frame}

@@

\bi
\item Two constructs isolated in the new scale, separately by cohort
\item But are externalising and internalising measured in a comparable way \emph{across cohorts}?
\ei
\vspace{12pt}\pause
\textbf{Measurement Invariance (MI)} analysis \citep{Wu2016a}
  \bi
  \item Formal test of the invariance hypothesis using factor analysis
  \item Aim: assess if constructs are measured in the same way in two or more groups
  \item Procedure:
    \bnm
    \item Start from the most general model -- parameters allowed differ across groups
    \item Impose restrictions on sets of parameters to be equal across groups
    \item Test relative fit of restricted models compared to initial model
    \enm
  \ei

@@

\begin{frame}
\frametitle{Measurement Invariance}\label{midet}
\framesubtitle{IFA model}
Categorical Item Factor Analysis (IFA) model

\bi
\item Threshold model \citep{Muthen1984}
\item Latent propensity to endorse each item $i$ for child $j$ in group $c$
\begin{align}\label{eq:propens}
\begin{split}
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
\end{split}
\end{align}

@@

We follow the approach in \cite{Wu2016a}
\bi
\item Define group ($c$) as cohort-gender cell
  \bi \item Invariance results do not vary if separating by gender \ei
\ei
~\\
\bnm
\item \textbf{Configural} model
  \bi
  \item Minimally restrictive (all parameters free across groups)
  \item Normalised on latent means $\bm{\kappa}$ and variances $\bm{\Phi}$
  \ei
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, and $\bm{\Lambda}$ to be equal across groups
  \item Allows comparison of latent factor variances
  \ei
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, $\bm{\Lambda}$, and $\bm{\nu}$ to be equal across groups
  \item Allows comparison of latent factor variances \emph{and} means
  \ei
\enm
\end{frame}

@@

\bi
\item We use RMSE to compare fit between models of different invariance level
\item Restricting thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit details}}
  \bi \item We can compare the scale (variance) of the latent factors \ei
\item Models with restricted intercepts do \emph{not} pass MI test
  \bi \item We cannot compare the location of latent factors in different groups \ei
\ei
\begin{center}
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

table <- rbind(afitab[[1]])
table <- as.matrix(table[,c("mnames","npar","rmsea")])

@@

print(xtable(table,
                    align="L{0cm}L{.5\\textwidth}C{.1\\textwidth}C{.15\\textwidth}",
                    digits = 4
                    ),
             size="\\footnotesize",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
            "Model & Num. params & RMSEA \\\\"
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
\end{center}
\end{frame}

\begin{frame}
\frametitle{Measurement Invariance}
\framesubtitle{Factor scores}
\bi
\item Nevertheless, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising dimensions)
\item Estimated parameters: \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}
\item Use Empirical Bayes approach to predict individual scores
\item All scores are coded so that \emph{higher is better}
\item \textbf{CAUTION}: location differences by cohort are not interpretable!
~\\
\pause
\item Is this an improvement on using simple sum scores?
  \bi
  \item Scores from the model exploit the covariance structure of the whole 11-item scale
  \item Every pattern of response is assigned a score
  \item Substantial increase in variability ({\color{red} could add boxplot of factor scores vs sum scores})
  \ei
\ei
\end{frame}

@@

@@

\section{Inequality in behaviour problems}

@@

\bi
\item \textbf{Q2}: Has socioeconomic inequality in behavioural problems increased between 1975 and 2005?
\item Investigate income gradient in externalising and internalising scores
  \bi
  \item Merge harmonised income data from CLOSER Work Package 2(a)
  \item Weekly family income when child is 10 (not contemporaneous)
  \item Working on access to harmonised social class data
  \ei
\ei

@@

\begin{frame}
\frametitle{Inequality}\label{ineq_uc}
\framesubtitle{Unconditional inequality}
\centering
<<INEQ_STATS, dependson="FA_SCORES_BIND", results='asis'>>=

@@

print(xtable(table,
                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
                    digits = 4
                    ),
             size="\\scriptsize",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,4)),
                              command = c(
                              # header
            "Measure & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
                       & BCS & MCS & BCS & MCS  \\\\
                      \\textbf{Externalising} \\\\ ",
                      "\\textbf{Internalising} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@@

\begin{frame}
\frametitle{Inequality}\label{ineq_sc}
\framesubtitle{Social class}
\vspace{-.2cm}
\begin{figure}
<<FACINEQ_SC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
# plot
fi <- fi_sc # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Inequality}\label{ineq_fsc}
\framesubtitle{Father occupation}
\vspace{-.2cm}
\begin{figure}
<<FACINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
# plot
fi <- fi_fsc # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Inequality}\label{ineq_ys}
\framesubtitle{Age mother left education}
\vspace{-.2cm}
\begin{figure}
<<FACINEQ_YS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
# plot
fi <- fi_ys # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\end{figure}
\end{frame}

@@

@@

@@

\begin{frame}
\frametitle{What explains behavioural skills?}

write about variables used and groups

@@

<<MCSEFA>>=
@

>>>>

% spacing
\usepackage{setspace}
\setstretch{1.3}
\setbeamerfont{frametitle}{size=\linespread{1.0}\Large}

@@

\author[Mason]{Orazio Attanasio \and Richard Blundell \\ Gabriella Conti \and Giacomo Mason  }
\institute[UCL \& IFS]{UCL and IFS}

@@

\begin{frame}
\frametitle{Motivation}
\framesubtitle{Inequality in skills}
\bi
\item Socio-emotional skills developed during childhood are important for lifetime outcomes
\item Socioeconomic gradients in these skills are persistent 
\item {\color{red} IMPROVE INTRO...}
\ei
\end{frame}

\begin{frame}
\frametitle{Motivation}
\framesubtitle{Measuring socioemotional skills}

\bi
  \item Usually, socioemotional constructs are measured using validated scales
    \bi \item e.g. Rutter, BSAG, SDQ \ei
  \item Most economic literature uses simple sum scores of these scales
  \item Sometimes useful to consider different populations/groups
    \bi
    \item Different countries
    \item Different subgroups (gender, race, ...)
    \item Same population, at different points in time
    \ei
  \item Do these scales measure the same constructs?
  \item Are the constructs measured in the same way?
  \item And what if only info from different scales is available?
\ei
\end{frame}

\begin{frame}
\frametitle{This study}
\framesubtitle{Research questions}
We compare the evolution of socioemotional skills in the UK
\bi
\item \textbf{Q1}: \textbf{How many latent dimensions} do we need to capture socioemotional skills in the data?
~\\
\pause
\item \textbf{Q2}: Can we derive \textbf{comparable measures} of socioemotional skills across cohorts of children born in different years?
~\\
\pause
\item \textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved in the UK between 1975 and 2005?
~\\
\pause
\item \textbf{Q4}: Have the \textbf{determinants} of socioemotional skills evolved?
~\\
\pause
\item \textbf{Q5}: How do early socioemotional skills map into \textbf{adolescent/adult outcomes}?
\ei
\end{frame}

@@

\bi
\item Focus on socioemotional skills in two British cohorts
  \bi \item British Cohort Study (BCS, b. 1970)
  \item Millennium Cohort Study (MCS, b. 2000/1) \ei
~\\ \pause
\item \textbf{Contribution}
\bi
\item Construct a common 11-item scale, starting from different survey instruments
\item Investigate its psychometric properties, and find
  \bnm 
  \item \textbf{\textit{Bidimensionality}} -- Externalising and Internalising
  \item \textbf{\textit{Comparability}} in terms of scale (but not location)
  \enm
\ei
~\\ \pause

\item \textbf{Key findings}
  \bi
  \item Inequality in non-cognitive skills has increased in 1975-2005
  \item Sharpest change for boys rather than girls
  \item {\color{red} OTHER RESULTS}
  \ei
\ei

@@

\bi
\item Four main cohort studies:
  \bi
  \item \textit{National Survey of Health and Development} (NSHD) -- b. March 1946
  \item \textit{National Child Development Study} (NCDS) -- b. March 1958
  \item \textit{British Cohort Study} (BCS) -- b. April 1970
  \item \textit{Millennium Cohort Study} (MCS) -- b. 2000-2001
  \ei
\item Nationally representative samples of 10,000+ individuals from birth
  \bi
  \item Perinatal period (pregnancy, birth outcomes)
  \item Skill formation -- cognitive, socioemotional
  \item Health and health behaviours
  \item Family background and composition
  \item Schooling and labour market outcomes
  \ei
\pause
\vspace{0.2cm}
\item Not always \textbf{consistent}
  \bi
  \item Different design, timing, questions, survey instruments...
  \item Harmonisation effort underway (CLOSER)
  \ei
\ei

@@

\begin{column}{0.6\textwidth}
\bi
\item Data from 5 year follow-up
  \bi \item Sample restricted to England \ei
\item Socioemotional skills: \textbf{Rutter A scale} \citep{Rutter1970}
  \bi
  \item 19 items (descriptions of behaviours) from maternal questionnaire
  \item Options: \textit{Doesn't apply, Applies somewhat, Certainly applies}
  \ei
  \item Additional questions from maternal module
\ei
\end{column}

@@

\bi
\item Data from 5 year follow-up
  \bi
  \item Sample restricted to England
  \item Complex survey design: rebalance sample to make representative
  \ei
\item Socioemotional skills: \textbf{Strengths and Difficulties Questionnaire} (SDQ, \citealp{Goodman1994})
  \bi
  \item ``Updated" and expanded version of the Rutter scale 
  \item 25 items (descriptions of behaviours)
  \item Options: \textit{Not true, Somewhat true, Certainly true}
  \ei
\ei

@@

\section{Deriving a harmonised scale}

@@

\bi
\item Rutter and SDQ are different instruments
  \bi \item Not ideal to just use sum scores and assume they return comparable information... \ei
\item Focus on subset of items that have similar wording between the two instruments
\item $\Rightarrow$ New 11-item subscale
\ei

@@

<<MEANTABLE_PRINT>>=
# function to print mean table
printmeantab <- function(tab) {
  print(xtable(tab,
               align="L{0cm}C{.04\\textwidth}L{.45\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}L{.45\\textwidth}C{.06\\textwidth}C{.06\\textwidth}C{.06\\textwidth}",
               digits = 1
  ), 
  size="\\small",
  include.rownames=F,
  include.colnames=F,
  booktabs = T,
  add.to.row = list(pos = c(list(0)),
                    command = c(
                      # header
                      paste(" & 
                                              \\multicolumn{4}{c}{\\textbf{BCS}} &
                                              \\multicolumn{4}{c}{\\textbf{MCS}} \\\\ \\cmidrule(lr){2-5} \\cmidrule(lr){6-9} 
                                            Item &
                                              Text & Cert. Appl. & Smt. Appl. & Appl. &
                                              Text & Cert. True & Smt. True & True \\\\ \\midrule
                                          "
                            , sep="")
                    )),
  sanitize.text.function = function(x){x},
  sanitize.rownames.function = function(x){""}, # to not print row numbers
  tabular.environment = "tabular", width = ".8\\textwidth",
  floating = FALSE
  )
}

\begin{frame}
\frametitle{Harmonisation}\label{meantab_f}
\framesubtitle{Choice of common items (females) \hyperlink{meantab_m}{\beamerbutton{Males}}}
\begin{spacing}{1.1}
\centering
\resizebox{!}{4cm}{
<<MEANTAB_F, dependson="MEANTABLE", results='asis'>>=
require(xtable)

@@

printmeantab(table)

@
} %scalebox
\end{spacing}
\end{frame}

@@

\textbf{Q1}: How many dimensions underlie the new 11-item scale?
\bi
\item Rutter A scale (BCS) 
  \bi 
  \item Traditionally conceptualised as bidimensional \citep{Blanden2007}
    \bi
    \item \emph{anti-social} -- or aggressive-hyperactive
    \item \emph{neurotic} -- or anxious-fearful
    \ei
  \item Sometimes the former is split into aggressive/conduct vs hyperkinetic \citep{McGee1985,Collishaw2004}
  \ei
\item Strengths and difficulties (MCS) 
  \bi
  \item Five subscales (emotional, hyperactivity, prosocial, peer, conduct) \citep{Stone2010}
  \item Evidence that lower-dimensional structures can work \citep{Goodman2010}
  \ei
\ei

@@

\bi
\item Different methods for assessing number of factors give varied results \hyperlink{numfac}{\beamerbutton{Det.}}
\item We choose two factors to maximise interpretability
\item In both cohorts, EFA delivers a clear-cut split between two factors
\item Factors can be coherently interpreted as externalising ($\sim$ anti-social) and internalising ($\sim$ neurotic)
\ei 
\vspace{12pt}
\centering
\begin{spacing}{1.1}
\begin{tabular}{C{.05\textwidth}L{.3\textwidth}|C{.05\textwidth}L{.3\textwidth}}
  & \textbf{{\color{OliveGreen}Externalising}}  &    & \textbf{{\color{Orange}Internalising}}  \\
1 & Restless                & 7  & Worried                \\
2 & Squirmy/fidgety         & 8  & Afraid of new things   \\
3 & Fights/bullies children & 9  & Solitary               \\
4 & Cannot settle           & 10 & Unhappy/tearful        \\
5 & Temper tantrums         & 11 & Head/stomachaches \\
6 & Disobedient             &    &                       
\end{tabular}
\end{spacing}

@@

\begin{frame}
\frametitle{EFA}\label{meantab_fcc2}
\framesubtitle{Choice of common items (females) \hyperlink{meantab_fcc3}{\beamerbutton{3 factors}}}
\begin{spacing}{1.1}
\centering
\resizebox{!}{4cm}{
<<MEANTAB_FCC, dependson="MEANTABLE", results='asis'>>=
require(xtable)

@@

printmeantab(table)

@
} %scalebox
\end{spacing}
\end{frame}

@@

\bi
\item Two constructs isolated in the new scale, separately by cohort
\item But are externalising and internalising measured in a comparable way \emph{across cohorts}?
\ei
\vspace{12pt}\pause
\textbf{Measurement Invariance (MI)} analysis \citep{Wu2016a}
  \bi
  \item Formal test of the invariance hypothesis using factor analysis
  \item Aim: assess if constructs are measured in the same way in 2+ groups
  \item Procedure:
    \bnm
    \item Start from the most general model -- parameters allowed differ across groups
    \item Impose restrictions on sets of parameters to be equal across groups
    \item Test relative fit of restricted models compared to initial model
    \enm
  \ei

@@

\begin{frame}
\frametitle{Measurement Invariance}\label{midet}
\framesubtitle{IFA model}
\setstretch{1.1}

Categorical Item Factor Analysis (IFA) model
\bi
\item Latent propensity to endorse each item $i$ for child $j$ in group $c$ \citep{Muthen1984}
\begin{align}\label{eq:propens}
\begin{split}
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{EXT}_{jc} + u_{ijc} \qquad \text{for} \; i=1,\dots, 6 \\
X^*_{ijc} &=\nu_{ci} + \lambda_{ci}\theta^{INT}_{jc} + u_{ijc} \qquad \text{for} \; i=7,\dots, 11
\end{split}
\end{align}

@@

We follow the approach in \cite{Wu2016a}
\bi
\item Define group ($c$) as cohort-gender cell
  \bi \item Invariance results do not vary if separating by gender \ei
\ei
\vspace{.2cm}
\bnm
\item \textbf{Configural} model
  \bi
  \item Minimally restrictive (all parameters free across groups)
  \item Normalised on latent means $\bm{\kappa}$ and variances $\bm{\Phi}$
  \ei
\item \textbf{Threshold and Loading invariance} model
  \bi
  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, and $\bm{\Lambda}$ to be equal across groups
  \item Allows comparison of latent factor variances
  \ei
\item \textbf{Threshold, Loading, and Intercept invariance} model
  \bi
  \item Restrict $\bm{\tau}_{1}$, $\bm{\tau}_{2}$, $\bm{\Lambda}$, and $\bm{\nu}$ to be equal across groups
  \item Allows comparison of latent factor variances \emph{and} means
  \ei
\enm
\end{frame}

@@

\bi
\item We use RMSE to compare fit between models of different invariance level
\item Restricting thresholds and loadings provides a good fit \hyperlink{fit}{\beamerbutton{Fit details}}
  \bi \item We can compare the scale (variance) of the latent factors \ei
\item Models with restricted intercepts do \emph{not} pass MI test
  \bi \item We cannot compare the location of latent factors in different groups \ei
\ei
\begin{center}
\resizebox{\textwidth}{!}{
<<FA_FIT_RES, dependson="FA_FIT", results='asis'>>=
require(xtable)

table <- rbind(afitab[[1]])
table <- merge(table,afitab[[2]], by="mnames") # merge additional result
table <- as.matrix(table[,c("mnames","npar.x","chisq.x","rmsea.x","chisq.y","rmsea.y")])

@@

print(xtable(table,
                    align="L{0cm}L{.35\\textwidth}C{.12\\textwidth}C{.15\\textwidth}C{.1\\textwidth}C{.15\\textwidth}C{.1\\textwidth}",
                    digits = 4
                    ),
             size="\\footnotesize",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0)),
                              command = c(
                              # header
            " & & \\multicolumn{2}{c}{\\textbf{Full sample}} & \\multicolumn{2}{c}{\\textbf{59-61 months}} \\\\ \\cmidrule(lr){3-4}  \\cmidrule(lr){5-6}
            Model & Num. par. & $\\chi^2$ & RMSEA & $\\chi^2$ & RMSEA \\\\"
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@
}
\end{center}
\end{frame}

@@

\begin{frame}
\frametitle{Measurement Invariance}
\framesubtitle{Factor scores}
\bi
\item Nevertheless, we can use the model with non-invariant intercepts to score the latent factors (externalising and internalising dimensions)
\item Estimated parameters: \hyperlink{measpars}{\beamerbutton{Measurement}}\hyperlink{lvpars}{\beamerbutton{Latent variable}}
\item Use Empirical Bayes approach to predict individual scores
\item All scores are coded so that \emph{higher is better}
\item \textbf{CAUTION}: location differences by cohort are not interpretable!
~\\
\pause
\item Is this an improvement on using simple sum scores?
  \bi
  \item Sum scores assume that each item conveys the same information
  \item Scores obtained from the model exploit the covariance structure of the whole 11-item scale
  \item Every 11-dimensional pattern of response in the data can be assigned a different score
  \item Substantial increase in variability ({\color{red} could add boxplot of factor scores vs sum scores})
  \ei
\ei
\end{frame}

@@

\section{Inequality in socioemotional skills}

@@

\bi
\item \textbf{Q3}: How has socioeconomic inequality in socioemotional skills increased between 1975 (BCS) and 2005 (MCS)?
\item 
\ei

@@

\begin{frame}
\frametitle{Inequality}\label{ineq_uc}
\framesubtitle{Unconditional (quantile differences)}
\centering
<<INEQ_STATS, dependson="FA_SCORES_BIND", results='asis'>>=

@@

print(xtable(table,
                    align="L{0cm}L{.15\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}C{.1\\textwidth}",
                    digits = 3
                    ),
             size="\\scriptsize",
             include.rownames=F,
             include.colnames=F,
             booktabs = T,
             add.to.row = list(pos = c(list(0,4)),
                              command = c(
                              # header
            "Measure & \\multicolumn{2}{c}{\\textbf{Males}} & \\multicolumn{2}{c}{\\textbf{Females}} \\\\ \\cmidrule(lr){2-3}  \\cmidrule(lr){4-5} 
                       & BCS & MCS & BCS & MCS  \\\\
                      \\textbf{Externalising} \\\\ ",
                      "\\textbf{Internalising} \\\\ \\midrule "
                                    )),
             sanitize.text.function = function(x){x},
             sanitize.rownames.function = function(x){""}, # to not print row numbers
             tabular.environment = "tabular", width = "\\textwidth",
             floating = FALSE
)

@@

\begin{frame}
\frametitle{Inequality}\label{ineq_fsc}
\framesubtitle{Father occupation \hyperlink{ineq_sc}{\beamerbutton{Social class (10)}}}
\vspace{-.2cm}
\begin{figure}
<<FACINEQ_FSC, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
# plot
fi <- fi_fsc # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Inequality}\label{ineq_ps}
\framesubtitle{Mother compulsory education \hyperlink{ineq_ys}{\beamerbutton{Age left education}}}
\vspace{-.2cm}
\begin{figure}
<<FACINEQ_PS, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=10, out.width='.9\\linewidth'>>=
# plot
fi <- fi_ps # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\end{figure}
\end{frame}

@@

\begin{frame}
\frametitle{Inequality}\label{ineq_sm}
\framesubtitle{Smoking in pregnancy}
\vspace{-.2cm}
\begin{figure}
<<FACINEQ_SM, dependson="FA_SCORES_BIND", dev='tikz', fig.height=7.5, fig.width=11, out.width='\\linewidth'>>=
# plot
fi <- fi_sm # select result
gridExtra::grid.arrange(fi$ineqlist[[1]],fi$ineqlist[[2]], 
                        fi$ineqlist[[3]],fi$ineqlist[[4]],
                        fi$histlist[[1]],fi$histlist[[2]], fi$legend,
                        ncol=2, nrow=4, heights = c(4,4,2,0.5))
@
\end{figure}
\end{frame}

\textbf{Q3}: How has \textbf{inequality} in early-life socioemotional skills evolved in the UK between 1975 and 2005?

\bi
\item Unconditionally, the distribution is more spread out (except for girls' internalising)
\item Inequality across socioeconomic indicators has increased, especially
  \bi
  \item for boys rather than girls
  \item considering inequality across maternal schooling and employment, rather than paternal occupation
  \ei
\ei

@@

\section{Determinants of socioemotional skills}

@@

\begin{frame}
\frametitle{Determinants of socioemotional skills}

We explain externalising and internalising skills using information from the birth and 5 year surveys \\

\begin{table}
\footnotesize
\begin{tabular}{L{.25\textwidth}C{.1\textwidth}L{.5\textwidth}}
Group & Age &  \\
\midrule
Father occupation & 5 & White collar (SC I-IIINM) \newline Blue collar (SC IIIM-V-other) \newline No father figure \\
Maternal education & 5 & Post-compulsory schooling \\
Maternal employment & 5 & Employed \\
Maternal background & 0/5 & Age, height \newline Nonwhite ethnicity (child) \newline Number of children in HH \\
Pregnancy & 0 & Firstborn child, number of previous stillbirths \newline Mother smoked in pregnancy \newline Preterm birth, (log) birth weight \\
\end{tabular}
\end{table}

@@

<<MCSEFA, dependson='LOAD_DATA'>>=
@
